Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language

### 2022-08-05 manuscript_v2.Rmd
updated files

Natural languages can be systematically used to keep track of previous referents in the discourse and mark the changing cognitive statuses of the REs (i.e., from being active to less active in the addressee's mind) to create a more coherent narrative. Among the theoretical frameworks that aim to explain that different linguistic forms are selected in accordance with the addressee's mental representations of the referents in the discourse is the Givenness Hierarchy [@gundel1993]. According to this model, given or familiar information is expressed in more attenuated forms such as pronominal use and null marking since the mental status of the expressed referent is already at the center of attention, as indicated by the present contextual information [@chafe1976]. The Givenness Hierarchy is akin to other previously proposed models of reference control within the topicality and saliency paradigm such as the Accessibility Hierarchy [@ariel1990], Topic Continuity [@givon1983], and Familiarity Scale [@prince1981]. @gundel1993 argued that the distribution of forms is not random but abides by the Gricean Maxim of Quantity [@grice1975]. That is to say, individuals must make the most informative and succinct choice possible to retain economy of form. When multiple expressions are available for the same referent as in an utterance like (1), phonetically simplest form is selected with the least redundant information are selected based on an economy principle. In (1), REs for "Eddie" are given in bold. All three REs (nominal, pronominal, and null marking) refer to the same entity; however, the quantity of marking decreases as the referent's accessibility increases. On this ground, for highly accessible referents (i.e., those that are highly active in the mind), linguistic forms with the least phonetic content are selected for the least amount of processing effort.

```{=tex}
\begin{exe}
    \ex
    So \textbf{Eddie} turned around. \hfill Williams (1988; p. 343, ex. 7) \\
\textbf{He} said, "youse got a problem?" \\
"Yeah, we want you," they say.
So-\textbf{$\emptyset$} walked right up to them \\
\end{exe}
```
It is also possible to categorize REs into three according to their discourse status [@gullberg2006]. A referent can be introduced or mentioned for the first time, maintained, that is, continued across two or more clauses, or re-introduced into the discourse following a topic shift. As a result, interlocutors mark some REs differently than others based on the principles of reference accessibility. For spoken languages, this observation is well established [@ariel1990; @givon1983; @gundel1993; @debreslioska2013; @ahrenholz2005; @chafe1976; @hickmann1996; @azar2015]. These research indicate that the introduction and re-introduction of a referent are often accomplished by full noun phrases which can be commonly modified by an adjective or demonstrative. Maintenance, on the other hand, is typically marked with pronominals and null markers since the continued referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken languages. There is an emerging body of literature that reports the distribution of REs in signed narratives. Building on @gundel1993's work, @swabey2002 demonstrated that, like other spoken languages such as Russian, Chinese, and Japanese, American Sign Language (ASL) signers use bare nouns for introduced and reintroduced discourse contexts whereas null marking is used for referents which are "in-focus" or very accessible to the addressee. Overt REs like pronouns or nouns are disprefered for such referents. @swabey2002 concluded that "native" signers in her study abided by the Gricean maxim of quantity and were thus economic in their reference tracking. Similar findings have been reported for German Sign Language (DGS) [@perniss2015] and British Sign Language (BSL) [@morgan2005]. It is important to note, however, that these research either collapsed native and late signers into a single group and labeled them as "native-like" or explicitly compared native deaf signers with late deaf signers (with language deprivation) in terms of narrative competence.

@@

Commonly in both spoken and sign languages nouns and noun phrases are used as referring expressions. Spoken languages use pronouns, determines and demonstratives where sign languages use a pointing sign, usually glossed as IX, which assumes similar functions [@lillo-martin1990]. In addition, the pointing sign is used to referentially anchor a referent to a location in signing space. For instance, when the signer introduces \textsc{elvan} for the first time, s/he can assign her to the right. This referential locus can later be used by agreement verbs and classifier predicates (more below). For instance, in (2a) there are two referents, \textsc{elvan}, who is referentially-assigned to the right side of the signer, represented with the "a" subscript and \textsc{demet}, who is referentially-assigned to the left side of the signer, represented with the "b" subscript. When the signer makes a contrast in (2b), as a pronominal use, she points to the right side to refer to ELVAN and to the left to refer to \textsc{demet}. We schematically represents the referential locus assigned to the referents in (3).

```{=tex}
\begin{exe}
    \ex 
\begin{xlist}
    \ex
    \textsc{lab place elvan ix\textsubscript{a} ten years work. demet ix\textsubscript{b} last year start.} \\
    ‘In the lab, Elvan has been working for 10 years. Demet started last year.’
    \\
    \ex
    \textsc{ix\textsubscript{a} e-l-a-n know. ix\textsubscript{b} learn.}\\
    ‘She (Elvan) knows ELAN (the annotation program). She (Demet) is learning.’
    \end{xlist}
    \end{exe}

    \begin{exe}
    \ex Signing space, signer, example referential loci for the referents \textsc{elvan} (a) and \textsc{demet} (b) \\

\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

@@

% Text Node
\draw (88.75,80.75) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\footnotesize Locus-b }};
% Text Node
\draw (91.75,90.25) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny (Demet)}};
% Text Node
\draw (208.25,80.75) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\footnotesize Locus-a }};
% Text Node
\draw (212.75,90.25) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny (Elvan)}};
% Text Node
\draw (144,115.25) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\footnotesize Signer}};
\end{tikzpicture}
    \end{exe}
```
Grammatical gender as in Romance languages or noun classes as in Bantu languages are some reference tracking tools that are familiar from spoken languages. A somehow less known reference-tracking tool in spoken languages is *verbal classifiers*, which morphologically attach to the verb and often cross-reference a core argument in a clause [@aikhenvald2017]. Classifiers denote sets of objects based on semantic criteria such as size, shape, body-part, texture and sometimes function. Usually, the verbal classifier and the cross-referenced noun co-occur in the same clause when the argument is introduced into discourse. The classifier can track the argument on its own on subsequent occurrences. In sign languages, a verbal classifier morpheme is encoded by a *handshape* that is articulated phonologically co-temporally with a movement, which encodes the verbal meaning [@supalla1990; @benedicto2004]. These two infused morphemes form a complex predicate. Some classifier handshapes and the set of objects that they denote are below. Some classifier handshapes and the set of objects that they denote are presented below in (4).

```{=tex}
\begin{exe}
\ex 
\includegraphics[width=25pt]{pictures/flathand.png}: flat entities, surfaces, vehicles.\\
\includegraphics[width=25pt]{pictures/flat.png}: human agent holding a flat entity.\\
\includegraphics[width=25pt]{pictures/small_legged.png}: small animal legs.\\
\includegraphics[width=25pt]{pictures/index.png}: thin entities (like a pencil); upright beings.\\
\includegraphics[width=25pt]{pictures/hold.png}: human agent holding a thin entity (like a pencil or a branch of flower). 
\end{exe}
```
In (5), in the first clause, a referential locus, i.e. locus-a, is assigned to the locative source argument \textsc{cage}. In the second clause, the core argument \textsc{bird} is cross-referenced with the classifier handshape, i.e. small-legged animal. It flies from the referential locus of the \textsc{cage} (locus-a) to the referential locus of the goal \textsc{tree} (locus-b).

```{=tex}
\begin{exe}
    \ex 
\begin{xlist}
    \ex \textsc{cage} assigned to the referential locus-a. \\ \\
    \textsc{garden bird cage-a open} \\
    ‘In the garden, there is a cage and it is open.’
    \\

@@

```{=tex}
\begin{multicols}{2}
\begin{exe}
\ex
\begin{xlist}
\ex Full noun \\\glll
\includegraphics[width=50pt]{pictures/1a_1.png}\\ \textsc{mouse\footnotemark} \\ \\ 
\ex Noun modified by classifier \\\glll
\includegraphics[width=50pt]{pictures/1c_2.png} \includegraphics[width=50pt]{pictures/1c_1.png} \\ \textsc{soap} \textsc{soap.cl} \\ \\ 
\ex Fingerspelled noun\\\glll 
\includegraphics[width=50pt]{pictures/1b_1.png} \includegraphics[width=50pt]{pictures/1b_2.png} \includegraphics[width=50pt]{pictures/1b_3.png}\\ \textsc{t-} \textsc{o-} \textsc{m} \\ \\
\ex Noun modified by index (IX)\\\glll 
\includegraphics[width=50pt]{pictures/1d_1.png} \includegraphics[width=50pt]{pictures/1d_2.png}\\ IX \textsc{cat} \\ \\
\end{xlist}
\end{exe}
\end{multicols}
\ea 
\ea Pronominal \\\glll
\includegraphics[width=50pt]{pictures/2_1.png} \includegraphics[width=50pt]{pictures/2_2.png} \\ IX \textsc{sick} \\ \\
\z \z
\begin{multicols}{4}
\begin{exe}
\ex
\begin{xlist}
\ex Whole entity \\\glll 
\includegraphics[width=50pt]{pictures/3e.png} \\ \textsc{fish.fall}\\ \\
\ex Body part \\\glll
\includegraphics[width=50pt]{pictures/3b.png} \\ \textsc{walk} \\ \\ 
\ex Handling \\\glll 
\includegraphics[width=50pt]{pictures/3c.png} \\ \textsc{hold} \\ \\
\ex Extension \\\glll 
\includegraphics[width=50pt]{pictures/3d_1.png} \includegraphics[width=50pt]{pictures/3a.png}\\ \textsc{plank} {\textsc{dh: mouse}} \\ {} {\textsc{ndh: tunnel}}\\ \\
\end{xlist}
\end{exe}
\end{multicols}
\begin{multicols}{3}
\begin{exe}
\ex
\begin{xlist}
\ex Constructed action \\\glll 
\includegraphics[width=50pt]{pictures/4c.png} \\ \textsc{struggle} \\ \\

\ex Plain verb  \\\glll
\includegraphics[width=50pt]{pictures/4a_2.png}  \\ \textsc{enter}\\ \\

\ex Agreement verb  \\\glll 
\includegraphics[width=50pt]{pictures/4b_1.png} 
\includegraphics[width=50pt]{pictures/4b_2.png} \\ {\textsc{$_{1}$hit$_{3}$}} {} \\ \\ 
\end{xlist}
\end{exe}
\end{multicols}
\footnotetext{We use small capitals to indicate sign glosses.}
```
### Nominals

In TİD and many other sign languages, it is possible to use bare noun phrases (\textsc{bare nom}) as in (1a) as well as fingerspelled nouns (\textsc{fs nom}) as in (1c). Fingerspelling refers to the use of the manual alphabet which is derived or borrowed from a surrounding written language [@kubu2008]. Nouns can also be modified (\textsc{mod nom}) by classifiers as in (1b) [@frederiksen2016] or by a prenominal or postnominal IX functioning as a determiner or a demonstrative as in (1d) [@maclaughlin1997].

@@

The IX sign (extended index finger) has various functions in sign languages, one of which is the pronominal use which refers to established nominals by pointing to a previously assigned abstract or physical locus in the signing space when re-introducing a topic in discourse [@emmorey1996]. The same IX function has also been identified for TİD [@sevinç2006]. The example in (2) illustrates a spatially-anchored pronominal reference for topic re-introduction. These signs are thus rarely used for the first mention of an entity [@perniss2015]. Other interpretations consider pointing signs without noun phrases as examples of demonstratives and not personal pronouns [@koulidobrova2016]. It is important to mention here that in some spoken languages (e.g., German and Dutch), demonstratives can also refer to persons [@bosch2007] and behave similarly to personal pronouns. However, for the purposes of the present study, we will refer to IX constructions as pronominal IX and particularly examine their non-deictic or personal pronoun realizations as a possible discourse strategy to mainly mark topic shift among native and late signers of TİD.

@@

Classifier predicates are highly complex morphological constructions which are very common in sign languages [@supalla1986]. These are manually signed iconic representations of animate or inanimate objects, and they embody several properties pertaining to the nominal referent including its location in the signing space, movement, path, size, and shape [@talmy2003]. As in most sign languages hitherto studied, classifier constructions are also common and found in TİD [see @dikyuva2015 for a review]. Following the classification of @engberg-pedersen1993, we distinguish between four types of classifier constructions. Whole entity classifiers (WECL) represent the entirety of any referent whereas body part classifiers (BPCL) only reflect a specific part of an animate entity (e.g., limbs, head, tongue, teeth, and so forth). Handling classifiers (HCL) occur with transitive verbs and depict how an object is handled or manipulated by agent. On this ground, an HCL[^1] depicts the manner of the referring action while the former two constructions encode the location of the entity as well as its motion in space. Extension classifiers (ExtCL), on the other hand, denote the size and shape of an entity by following a trajectory to trace its shape. We exemplify these constructions in (3).

@@

Parallel to spoken languages, many sign languages have been reported to allow null subjects and/or objects and make use of null arguments when referring to highly accessible referents which are still active in the mental representation of the addressee. In signed discourse, this can be achieved by employing certain verbal constructions. To identify the predicate types in ASL, @padden1986 suggested a triadic categorization of verbs in sign languages: plain verbs, agreement verbs, and spatial verbs. According to this classification, plain verbs (plain) (4b) do not agree with the subject or the object. This means that they are not inflected for number or person. This contrasts with agreeing verbs (agreement) (4c) the starting or the ending point of which can be modified to specific loci for person and number agreement while spatial verbs (as in \textsc{ix book put} 'I placed the book there') agree with locative arguments but like plain verbs, they are not inflected for number or person. TİD too licenses null arguments. Following this classification, @sevinç2006 argued that although it is common in TİD to drop arguments, there were certain limitations on the null marking of the arguments for agreement and plain verbs.

@@

In both communication modalities (visual and auditory), children start to produce more complex and coherent stories with appropriate and informative linguistic forms to indicate anaphoric or cataphoric references after age 10 [@hickmann1996; @rathmann2007]. Prior to this age, both signing and speaking children fail to refer to entities consistently since their pragmatic abilities are yet to develop. The manipulation of linguistic forms for discourse functions to signal topic continuance or shift is difficult to master until the beginning of formal education. These findings are also comparable across different sign languages including British Sign Language (BSL) [@morgan2005], ASL [@loew1984], and German Sign Language (Deutsche Gebärdensprache - DGS) [@becker2009], pointing out to the fact that the mastery of reference tracking in narratives is one of the most protracted achievements in the acquisition of signed discourse.

@@

Still, only very few research examined whether linguistically deprived signers (i.e., deaf signers with delayed exposure to a sign language) also display redundancy in their reference tracking. @becker2009 examined 6 DGS signing deaf children (3 native and 3 late)'s narrative abilities. She reported that late signing children generated more explicit forms such as nominals when referring to the characters in the story compared to native learners, concluding that first language deprivation results in an inadequate representation of the addressee and less efficient use of morphosyntax. Comparably, @cormier2013 investigated the use of constructed action in 15 BSL among deaf native and late adult signers and found that late signers were redundant in topic continuity. In other words, the late signers produced more explicit forms like nominals, and fewer instances of null marking when maintaining previous referents. Most relevantly to the present study, @gür2018 studied the development of narrative skills among 10 native and 10 late deaf children and deaf adult signers of TİD. She found that native-signing children and adults had more instances of explicit referent introduction, set the scene more frequently, and employed constructed action in an appropriate manner more than corresponding late-signing children and adults. Although this study presented preliminary findings on age of acquisition effects on narrative competence, it does not investigate how native and late signers employ different linguistic tools based on the accessibility of the referent. Therefore, to our knowledge, there is a lack of research directly examining the effects of first language deprivation on deaf adult signers' selection of REs in signed narratives.

@@

In an attempt at addressing this gap in the literature, the present study aims to investigate the reference tracking strategies of native and late adult signers in Turkish Sign Language (Türk İşaret Dili - TİD) narratives with a story-telling paradigm. We explore the forms available for referent tracking in TİD and examine the nature of RE selection as a possible indicator of pragmatic competence between our acquisition groups.

@@

Participants sat in front of a laptop computer which had the stimulus stories. The instructions were all given by a deaf research assistant who was present in all sessions and ensured that the task requirements were clear. The informed consent form was translated into TİD and explained to the participants by the same assistant before data recording. Participants filled in the consent form and a background questionnaire form before the onset of the trials. Following this step, signers were seated across a hearing researcher who is an advanced late learner of TİD. We instructed the participants to narrate as they would to a deaf friend. All trials were recorded by a digital camera placed in front of the signers. The participants were instructed to watch the narrative clip as many times as they needed, and narrate the content of each clip either to the experimenter or the recording digital camera\footnote{Only 4 participants (2 native, 2 late) narrated to the experimenter whereas the remaining e did not make it obligatory for participants to either narrate to the digital camera or to the experimenter for few reasons. First, a native or native-like deaf experimenter was not always available, and a hearing late signer of TİD attended some sessions as the experimenter. Therefore, most signers preferred to narrate to the camera. Furhermore, some signers (4 in number) did not feel comfortable narrating to the camera and, as a result, narrated to the deaf research assistant, if available. We ran an additional model excluding the 4 signers who narrated to the experimenter and compared it with the original model including these signers in Appendix. The results imply that there is not much of a difference.}.

@@

[^2]: We ran another model where we used the Nativeness predictor with 3 levels 0-3:Native, 4-7:Late1, and 7-above:Late2. However, our model did not converg properly. We provide the model results as a plot and table in the Appendix.

@@

# Effect of narrating to the camera vs. experimenter
\label{section:b}
```{r echo=FALSE, message=FALSE, cache=FALSE, results='asis', warnings=FALSE}

>>>>

Natural languages can be systematically used to keep track of previous referents in the discourse and mark the changing cognitive statuses of the REs (i.e., from being active to less active in the addressee's mind) to create a more coherent narrative. Among the theoretical frameworks that aim to explain that different linguistic forms are selected in accordance with the addressee's mental representations of the referents in the discourse is the Givenness Hierarchy [@gundel1993]. According to this model, given or familiar information is expressed in more attenuated forms such as pronominal use and null marking since the mental status of the expressed referent is already at the center of attention, as indicated by the present contextual information [@chafe1976]. The Givenness Hierarchy is akin to other previously proposed models of reference control within the topicality and saliency paradigm such as the Accessibility Hierarchy [@ariel1990], Topic Continuity [@givon1983], and Familiarity Scale [@prince1981]. @gundel1993 argued that the distribution of forms is not random but abides by the Gricean Maxim of Quantity [@grice1975]. That is to say, individuals must make the most informative and succinct choice possible to retain economy of form. When multiple expressions are available for the same referent as in an utterance like (\ref{ex:1}), phonetically simplest form is selected with the least redundant information are selected based on an economy principle. In the utterance, REs for "Eddie" are given in bold. All three REs (nominal, pronominal, and null marking) refer to the same entity; however, the quantity of marking decreases as the referent's accessibility increases. On this ground, for highly accessible referents (i.e., those that are highly active in the mind), linguistic forms with the least phonetic content are selected for the least amount of processing effort.

```{=tex}
\begin{exe}
    \ex \label{ex:1}
    So \textbf{Eddie} turned around. \hfill Williams (1988; p. 343, ex. 7) \\
\textbf{He} said, "youse got a problem?" \\
"Yeah, we want you," they say.
So-$\pmb{\emptyset}$ walked right up to them \\
\end{exe}
```
It is also possible to categorize REs into three according to their discourse status [@gullberg2006]. A referent can be introduced or mentioned for the first time, maintained, that is, continued across two or more clauses, or re-introduced into the discourse following a topic shift. As a result, interlocutors mark some REs differently than others based on the principles of reference accessibility. For spoken languages, this observation is well established [@ariel1990; @givon1983; @gundel1993; @debreslioska2013; @ahrenholz2005; @chafe1976; @hickmann1996; @azar2015]. These research indicate that the introduction and re-introduction of a referent are often accomplished by full noun phrases which can be commonly modified by an adjective or demonstrative. Maintenance, on the other hand, is typically marked with pronominals and null markers since the continued referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken languages. There is an emerging body of literature that reports the distribution of REs in signed narratives. Building on @gundel1993's work, @swabey2002 demonstrated that, like other spoken languages such as Russian, Chinese, and Japanese, American Sign Language (ASL) signers use bare nouns for introduced and re-introduced discourse contexts whereas null marking is used for referents which are "in-focus" or very accessible to the addressee. Overt REs like pronouns or nouns are disprefered for such referents. @swabey2002 concluded that "native" signers in her study abided by the Gricean maxim of quantity and were thus economic in their reference tracking. Similar findings have been reported for German Sign Language (Deutsche Gebärdensprache - DGS) [@perniss2015] and British Sign Language (BSL) [@morgan2005]. It is important to note, however, that these research either collapsed native and late signers into a single group and labeled them as "native-like" or explicitly compared native deaf signers with late deaf signers (with language deprivation) in terms of narrative competence.

@@

Commonly in both spoken and sign languages nouns and noun phrases are used as REs. Spoken languages use pronouns, determines and demonstratives where sign languages use a pointing sign, usually glossed as IX, which assumes similar functions [@lillo-martin1990]. In addition, the pointing sign is used to referentially anchor a referent to a location in signing space. For instance, when the signer introduces \textsc{elvan} for the first time, s/he can assign her to the right. This referential locus can later be used by agreement verbs and classifier predicates (more below). For instance, in (\ref{ex:2a}) there are two referents, \textsc{elvan}, who is referentially-assigned to the right side of the signer, represented with the "a" subscript and \textsc{demet}, who is referentially-assigned to the left side of the signer, represented with the "b" subscript. When the signer makes a contrast in (\ref{ex:2b}), as a pronominal use, she points to the right side to refer to ELVAN and to the left to refer to \textsc{demet}. We schematically represents the referential locus assigned to the referents in (\ref{ex:3}).

```{=tex}
\begin{exe}
    \ex 
\begin{xlist}
    \ex \label{ex:2a}
    \textsc{lab place elvan ix\textsubscript{a} ten years work. demet ix\textsubscript{b} last year start.} \\
    ‘In the lab, Elvan has been working for 10 years. Demet started last year.’
    \\
    \ex \label{ex:2b}
    \textsc{ix\textsubscript{a} e-l-a-n know. ix\textsubscript{b} learn.}\\
    ‘She (Elvan) knows ELAN (the annotation program). She (Demet) is learning.’
    \end{xlist}
    \end{exe}

    \begin{exe}
    \ex \label{ex:3} Signing space, signer, example referential loci for the referents \textsc{elvan} (a) and \textsc{demet} (b) \\
\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

@@

% Text Node
\draw (88.75,80.75) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\footnotesize Locus-b }};
% Text Node
\draw (91.75,90.25) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny (Demet)}};
% Text Node
\draw (208.25,80.75) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\footnotesize Locus-a }};
% Text Node
\draw (212.75,90.25) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\tiny (Elvan)}};
% Text Node
\draw (144,115.25) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\footnotesize Signer}};
\end{tikzpicture}
    \end{exe}
```
Grammatical gender as in Romance languages or noun classes as in Bantu languages are some reference tracking tools that are familiar from spoken languages. A somehow less known reference-tracking tool in spoken languages is *verbal classifiers*, which morphologically attach to the verb and often cross-reference a core argument in a clause [@aikhenvald2017]. Classifiers denote sets of objects based on semantic criteria such as size, shape, body-part, texture and sometimes function. Usually, the verbal classifier and the cross-referenced noun co-occur in the same clause when the argument is introduced into discourse. The classifier can track the argument on its own on subsequent occurrences. In sign languages, a verbal classifier morpheme is encoded by a *handshape* that is articulated phonologically co-temporally with a movement, which encodes the verbal meaning [@supalla1990; @benedicto2004]. These two infused morphemes form a complex predicate. Some classifier handshapes and the set of objects that they denote are below. Some classifier handshapes and the set of objects that they denote are presented below in (\ref{ex:4}).

```{=tex}
\begin{exe}
\ex \label{ex:4}
\includegraphics[width=25pt]{pictures/flathand.png}: flat entities, surfaces, vehicles.\\
\includegraphics[width=25pt]{pictures/flat.png}: human agent holding a flat entity.\\
\includegraphics[width=25pt]{pictures/small_legged.png}: small animal legs.\\
\includegraphics[width=25pt]{pictures/index.png}: thin entities (like a pencil); upright beings.\\
\includegraphics[width=25pt]{pictures/hold.png}: human agent holding a thin entity (like a pencil or a branch of flower). 
\end{exe}
```
In (\ref{ex:5}), in the first clause, a referential locus, i.e. locus-a, is assigned to the locative source argument \textsc{cage}. In the second clause, the core argument \textsc{bird} is cross-referenced with the classifier handshape, i.e. small-legged animal. It flies from the referential locus of the \textsc{cage} (locus-a) to the referential locus of the goal \textsc{tree} (locus-b).

```{=tex}
\begin{exe}
    \ex \label{ex:5}
\begin{xlist}
    \ex \textsc{cage} assigned to the referential locus-a. \\ \\
    \textsc{garden bird cage-a open} \\
    ‘In the garden, there is a cage and it is open.’
    \\

@@

```{=tex}
\begin{multicols}{2}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:6a} Full noun \\\glll
\includegraphics[width=50pt]{pictures/1a_1.png}\\ \textsc{mouse\footnotemark} \\ \\ 
\ex \label{ex:6b} Noun modified by classifier \\\glll
\includegraphics[width=50pt]{pictures/1c_2.png} \includegraphics[width=50pt]{pictures/1c_1.png} \\ \textsc{soap} \textsc{soap.cl} \\ \\ 
\ex \label{ex:6c} Fingerspelled noun\\\glll 
\includegraphics[width=50pt]{pictures/1b_1.png} \includegraphics[width=50pt]{pictures/1b_2.png} \includegraphics[width=50pt]{pictures/1b_3.png}\\ \textsc{t-} \textsc{o-} \textsc{m} \\ \\
\ex \label{ex:6d} Noun modified by index (\textsc{IX})\\\glll 
\includegraphics[width=50pt]{pictures/1d_1.png} \includegraphics[width=50pt]{pictures/1d_2.png}\\ \textsc{ix} \textsc{cat} \\ \\
\end{xlist}
\end{exe}
\end{multicols}
\ea 
\ea \label{ex:7} Pronominal \\\glll
\includegraphics[width=50pt]{pictures/2_1.png} \includegraphics[width=50pt]{pictures/2_2.png} \\ \textsc{ix} \textsc{sick} \\ \\
\z \z
\begin{multicols}{4}
\begin{exe}
\ex \label{ex:8}
\begin{xlist}
\ex Whole entity \\\glll 
\includegraphics[width=50pt]{pictures/3e.png} \\ \textsc{fish.fall}\\ \\
\ex Body part \\\glll
\includegraphics[width=50pt]{pictures/3b.png} \\ \textsc{walk} \\ \\ 
\ex Handling \\\glll 
\includegraphics[width=50pt]{pictures/3c.png} \\ \textsc{hold} \\ \\
\ex Extension \\\glll 
\includegraphics[width=50pt]{pictures/3d_1.png} \includegraphics[width=50pt]{pictures/3a.png}\\ \textsc{plank} {\textsc{dh: mouse}} \\ {} {\textsc{ndh: tunnel}}\\ \\
\end{xlist}
\end{exe}
\end{multicols}
\begin{multicols}{3}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:9a} Constructed action \\\glll 
\includegraphics[width=50pt]{pictures/4c.png} \\ \textsc{struggle} \\ \\

\ex \label{ex:9b} Plain verb  \\\glll
\includegraphics[width=50pt]{pictures/4a_2.png}  \\ \textsc{enter}\\ \\

\ex \label{ex:9c} Agreement verb  \\\glll 
\includegraphics[width=50pt]{pictures/4b_1.png} 
\includegraphics[width=50pt]{pictures/4b_2.png} \\ {\textsc{$_{1}$hit$_{3}$}} {} \\ \\ 
\end{xlist}
\end{exe}
\end{multicols}
\footnotetext{We use small capitals to indicate sign glosses.}
```
### Nominals

In TİD and many other sign languages, it is possible to use bare noun phrases as in (\ref{ex:6a}) as well as fingerspelled nouns as in (\ref{ex:6c}). Fingerspelling refers to the use of the manual alphabet which is derived or borrowed from a surrounding written language [@kubu2008]. Nouns can also be modified by classifiers as in (\ref{ex:6b}) [@frederiksen2016] or by a prenominal or postnominal IX functioning as a determiner or a demonstrative as in (\ref{ex:6d}) [@maclaughlin1997].

@@

The IX sign (extended index finger) has various functions in sign languages, one of which is the pronominal use which refers to established nominals by pointing to a previously assigned abstract or physical locus in the signing space when re-introducing a topic in discourse [@emmorey1996]. The same IX function has also been identified for TİD [@sevinç2006]. The example in (\ref{ex:7}) illustrates a spatially-anchored pronominal reference for topic re-introduction. These signs are thus rarely used for the first mention of an entity [@perniss2015]. Other interpretations consider pointing signs without noun phrases as examples of demonstratives and not personal pronouns [@koulidobrova2016]. It is important to mention here that in some spoken languages (e.g., German and Dutch), demonstratives can also refer to persons [@bosch2007] and behave similarly to personal pronouns. However, for the purposes of the present study, we will refer to IX constructions as pronominal IX and particularly examine their non-deictic or personal pronoun realizations as a possible discourse strategy to mainly mark topic shift among native and late signers of TİD.

@@

Classifier predicates are highly complex morphological constructions which are very common in sign languages [@supalla1986]. These are manually signed iconic representations of animate or inanimate objects, and they embody several properties pertaining to the nominal referent including its location in the signing space, movement, path, size, and shape [@talmy2003]. As in most sign languages hitherto studied, classifier constructions are also common and found in TİD [see @dikyuva2015 for a review]. Following the classification of @engberg-pedersen1993, we distinguish between four types of classifier constructions. Whole entity classifiers (WECL) represent the entirety of any referent whereas body part classifiers (BPCL) only reflect a specific part of an animate entity (e.g., limbs, head, tongue, teeth, and so forth). Handling classifiers (HCL) occur with transitive verbs and depict how an object is handled or manipulated by agent. On this ground, an HCL[^1] depicts the manner of the referring action while the former two constructions encode the location of the entity as well as its motion in space. Extension classifiers (ExtCL), on the other hand, denote the size and shape of an entity by following a trajectory to trace its shape. We exemplify these constructions in (\ref{ex:8}).

@@

Parallel to spoken languages, many sign languages have been reported to allow null subjects and/or objects and make use of null arguments when referring to highly accessible referents which are still active in the mental representation of the addressee. In signed discourse, this can be achieved by employing certain verbal constructions or constructed action (CA). The latter construction exemplified in (\ref{ex:9a}) refers to cases in which the signers engage with the space in front of themselves as if they are interacting with the objects and characters in the story. This is often achieved with the signer taking on the role of a referent in the narrative and performing the actions [@metzger1995]. CA is often marked by certain non-manual cues such as the facial expression and the breaking of the eye contact. This is to say, the narrator's head, face, torso, arms, and as an optional marker, hands represent those of the associated referent(s) [@smith2014]. As for the verbal constructions, to identify the predicate types in ASL, @padden1986 suggested a triadic categorization of verbs in sign languages: plain verbs, agreement verbs, and spatial verbs. According to this classification, plain verbs (\ref{ex:9b}) do not agree with the subject or the object. This means that they are not inflected for number or person. This contrasts with agreeing verbs (\ref{ex:9c}) the starting or the ending point of which can be modified to specific loci for person and number agreement while spatial verbs (as in \textsc{ix book put} 'I placed the book there') agree with locative arguments but like plain verbs, they are not inflected for number or person. TİD too licenses null arguments. Following this classification, @sevinç2006 argued that although it is common in TİD to drop arguments, there were certain limitations on the null marking of the arguments for agreement and plain verbs.

@@

In both communication modalities (visual and auditory), children start to produce more complex and coherent stories with appropriate and informative linguistic forms to indicate anaphoric or cataphoric references after age 10 [@hickmann1996; @rathmann2007]. Prior to this age, both signing and speaking children fail to refer to entities consistently since their pragmatic abilities are yet to develop. The manipulation of linguistic forms for discourse functions to signal topic continuance or shift is difficult to master until the beginning of formal education. These findings are also comparable across different sign languages including BSL [@morgan2005], ASL [@loew1984], and DGS [@becker2009], pointing out to the fact that the mastery of reference tracking in narratives is one of the most protracted achievements in the acquisition of signed discourse.

@@

Still, only very few research examined whether linguistically deprived signers (i.e., deaf signers with delayed exposure to a sign language) also display redundancy in their reference tracking. @becker2009 examined 6 DGS signing deaf children (3 native and 3 late)'s narrative abilities. She reported that late signing children generated more explicit forms such as nominals when referring to the characters in the story compared to native learners, concluding that first language deprivation results in an inadequate representation of the addressee and less efficient use of morphosyntax. Comparably, @cormier2013 investigated the use of CA in 15 BSL among deaf native and late adult signers and found that late signers were redundant in topic continuity. In other words, the late signers produced more explicit forms like nominals, and fewer instances of null marking when maintaining previous referents. Most relevantly to the present study, @gür2018 studied the development of narrative skills among 10 native and 10 late deaf children and deaf adult signers of TİD. She found that native-signing children and adults had more instances of explicit referent introduction, set the scene more frequently, and employed CA in an appropriate manner more than corresponding late-signing children and adults. Although this study presented preliminary findings on age of acquisition effects on narrative competence, it does not investigate how native and late signers employ different linguistic tools based on the accessibility of the referent. Therefore, to our knowledge, there is a lack of research directly examining the effects of first language deprivation on deaf adult signers' selection of REs in signed narratives.

@@

In an attempt at addressing this gap in the literature, the present study aims to investigate the reference tracking strategies of native and late adult signers in TİD narratives with a story-telling paradigm. We explore the forms available for referent tracking in TİD and examine the nature of RE selection as a possible indicator of pragmatic competence between our acquisition groups.

@@

Participants sat in front of a laptop computer which had the stimulus stories. The instructions were all given by a deaf research assistant who was present in all sessions and ensured that the task requirements were clear. The informed consent form was translated into TİD and explained to the participants by the same assistant before data recording. Participants filled in the consent form and a background questionnaire form before the onset of the trials. Following this step, signers were seated across a hearing researcher who is an advanced late learner of TİD. We instructed the participants to narrate as they would to a deaf friend. All trials were recorded by a digital camera placed in front of the signers. The participants were instructed to watch the narrative clip as many times as they needed, and narrate the content of each clip either to the experimenter or the recording digital camera\footnote{Only 4 participants (2 native, 2 late) narrated to the experimenter whereas the remaining e did not make it obligatory for participants to either narrate to the digital camera or to the experimenter for few reasons. First, a native or native-like deaf experimenter was not always available, and a hearing late signer of TİD attended some sessions as the experimenter. Therefore, most signers preferred to narrate to the camera. Furhermore, some signers did not feel comfortable narrating to the camera and, as a result, narrated to the deaf research assistant, if available. We ran an additional model excluding the 4 signers who narrated to the experimenter and compared it with the original model including these signers (see \ref{section:c} The results imply that there is not much of a difference.}.

@@

[^2]: We ran another model where we used the Nativeness predictor with 3 levels 0-3:Native, 4-7:Late1, and 7-above:Late2. However, our model did not converg properly. We provide the model results as a plot and table in the \ref{section:d}.

@@

```{=tex}
\setlength{\columnsep}{10pt}
\twocolumn
\linespread{1}
\justifying
\footnotesize
```

# Episodes in the stimuli
